---
title: "TCR Repertoire Diversity and Clonality Analysis"
author: "Avital Abergel Klein"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(immunarch)
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```


This report analyzes TCR repertoire diversity and clonality across blood and tumor samples in different experimental groups and time points.

# Functions

```{r functions}
# Process by group
process_by_group <- function(immdata, group_name) {
  filtered <- repFilter(immdata, .method = "by.meta", .query = list(class = include(group_name)))
  filtered$data <- repSample(filtered$data, "downsample")

  diversity_measures <- list(
    chao1 = repDiversity(filtered$data, "chao1"),
    hill = repDiversity(filtered$data, "hill"),
    d50 = repDiversity(filtered$data, "d50"),
    div = repDiversity(filtered$data, "div"),
    gini_simp = repDiversity(filtered$data, "gini.simp"),
    inv_simp = repDiversity(filtered$data, "inv.simp")
  )
  diversity_plots <- lapply(diversity_measures, function(div) vis(div, .by = "tp", .meta = filtered$meta))

  return(list(
    data = filtered$data,
    plots = diversity_plots
  ))
}

# Process by time point
process_by_timepoint <- function(immdata, timepoint) {
  filtered <- repFilter(immdata, .method = "by.meta", .query = list(tp = include(timepoint)))
  filtered$data <- repSample(filtered$data, "downsample")

  diversity_measures <- list(
    chao1 = repDiversity(filtered$data, "chao1"),
    hill = repDiversity(filtered$data, "hill"),
    d50 = repDiversity(filtered$data, "d50"),
    div = repDiversity(filtered$data, "div"),
    gini_simp = repDiversity(filtered$data, "gini.simp"),
    inv_simp = repDiversity(filtered$data, "inv.simp")
  )
  diversity_plots <- lapply(diversity_measures, function(div) vis(div, .by = "class", .meta = filtered$meta))

  return(list(
    data = filtered$data,
    plots = diversity_plots
  ))
}
```

# Load Data

```{r load-data}
# Load TCR sequencing data
immdata_AB <- readRDS("../../GBM/rawData_trimmed_200bp_correct/TRA_TRB/immdata.RDS")
```

# Metadata Preparation

```{r metadata-preparation}
metadata_AB <- immdata_AB$meta %>%
  mutate(tp = factor(tp, levels = c("0", "7", "21", "35", "49", "63", "Tumor")))
```

# Exploratory Analysis: Volume and Clone Counts

```{r exploratory-analysis}
# Fig.6. Exploratory visualization before filtering Tumor samples.
exp_vol <- repExplore(immdata_AB$data, .method = "volume")
exp_cln <- repExplore(immdata_AB$data, .method = "clone")

# Fig.6.a : Clonotype counts per sample by treatment group.
vis(volume_explore, .by = "class", .meta = immdata_AB$meta) + theme_classic()
# Fig.6.b : Clone counts per sample by treatment group.
vis(clone_explore, .by = "class", .meta = immdata_AB$meta) + theme_classic()
# Fig.6.c : Clonotype counts per sample by time point.
vis(volume_explore, .by = "tp", .meta = immdata_AB$meta) + theme_classic()
# Fig.6.d : Clone counts per sample by time point. 
vis(clone_explore, .by = "tp", .meta = immdata_AB$meta) + theme_classic()
```

# Filter Out Tumor Samples

```{r filtering-tumor}
immdata <- repFilter(immdata_AB, .method = "by.meta", .query = list(tp = exclude("Tumor")))
```

# Diversity and Clonality Analysis by Group

```{r group-analysis, warning=FALSE}
#responders_results <- process_by_group(immdata, "Responders")
#nonresponders_results <- process_by_group(immdata, "Non-responders")
#control_results <- process_by_group(immdata, "Control")
```

# Diversity and Clonality Analysis by Time Point

```{r timepoint-analysis, warning=FALSE}
#t0_results <- process_by_timepoint(immdata, "0")
#t7_results <- process_by_timepoint(immdata, "7")
#t21_results <- process_by_timepoint(immdata, "21")
```

# Clonality Analysis: Top and Rare Clones

```{r clonality-analysis}
# Downsample for fair comparison
immdata$data <- repSample(immdata$data, "downsample")

# Top and rare clonality
top_clonality <- repClonality(immdata$data, .method = "top", .head = c(10, 100, 1000, 3000, 10000))
rare_clonality <- repClonality(immdata$data, .method = "rare")
# Fig.7. Measurement of clonal proportions across blood samples following downsampling.
vis(top_clonality, .by = "class", .meta = immdata$meta) + theme_classic()
vis(rare_clonality, .by = "class", .meta = immdata$meta) + theme_classic()
```

```{r diversity-analysis, warning=FALSE}
# Fig.8. TCR repertoire diversity
div_chao <- repDiversity(immdata$data, "chao1")
div_hill <- repDiversity(immdata$data, "hill")
div_d50 <- repDiversity(immdata$data, "d50")
div_div <- repDiversity(immdata$data, "div")
div_ginisimp <- repDiversity(immdata$data, "gini.simp")
div_invsimp <- repDiversity(immdata$data, "inv.simp")
div_gini <- repDiversity(immdata$data, "gini")


vis(div_div, .by = c("class"), .meta = immdata$meta) + theme_classic() + theme(legend.position = "none")
vis(div_invsimp, .by = c("class"), .meta = immdata$meta) + theme_classic() + theme(legend.position = "none")

vis(div_div, .by = c("tp"), .meta = immdata$meta) + theme_classic() + theme(legend.position = "none")
vis(div_invsimp, .by = c("tp"), .meta = immdata$meta) + theme_classic() + theme(legend.position = "none")

```

